---
title: "Group8_Project2"
author: "Daniel Hostadt"
date: "2023-11-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1
# b)
```{r}
rm(list = ls())
```

Determining the parameters of the queueing system.
```{r}
lambda = 5
mu = 6  # 1/mu = 10 minutes = 1/6 hour, so mu = 6
length = 50*24  # 50 days à 24 hours
```

Preparing the initial conditions for the time and the state of the queue.
```{r}
state = c(0)
time = c(0)
```

Simulate the birth-and-death process as shown in the queue sample code.
```{r}
set.seed(1)
# As long as the last element of the time vector does not exceed the length
while (tail(time,1) < length) {
  current_state = tail(state,1)
  # Cases if currently there are zero patients or at least one patient
  # Then, append the state and the time stamp of the process
  if (current_state == 0) {
    time = c(time, tail(time,1) + rexp(n=1, rate=lambda))
    state = c(state, 1)
  } else {
    time = c(time, tail(time,1) + rexp(n=1, rate=lambda+mu))
    if (runif(1) < lambda/(lambda+mu)) {
      state = c(state, current_state+1)
    } else {
      state = c(state, current_state-1)
    }
  }
}
```

Calculate the expected time spent in each state.
```{r}
# Create the mass distribution
state_table = rep(0, 81)
for (i in 1:(length(time)-1)) {
  state_table[state[i]+1] = state_table[state[i]+1] + time[i+1] - time[i]
}

# Calculate the expected number of patients in the UCC
pi = state_table/time[length(state)-1]
L = sum((0:80)*pi)

# Calculate the expected waiting time
W = L/lambda
print(W)
```

Plot one realization of the time series for 0-12 hours.
```{r}
plot(NULL, NULL, xlim = c(0, 12), ylim = c(0, 10),
     xlab = "Time (hours)", ylab = "Number of patients in the UCC",
     main = "Simulation of number of patients in the UCC",
     cex.axis = 1.5, cex.lab = 1.5)
# As long as the time stamp of 12 hours is not exceeded
for (i in 1:sum(time < 12)) {
  lines(time[i:(i+1)], rep(state[i], 2), lwd = 2)
}
```

Repeat the simulation 30 times to construct a confidence interval.
```{r}
# For each simulation, write the expected waiting time to the vector
waiting_times = rep(0, 30)
set.seed(1)
for (i in 1:30) {
  state = c(0)
  time = c(0)
  
  while (tail(time,1) < length) {
    current_state = tail(state,1)
    if (current_state == 0) {
      time = c(time, tail(time,1) + rexp(n=1, rate=lambda))
      state = c(state, 1)
    } else {
      time = c(time, tail(time,1) + rexp(n=1, rate=lambda+mu))
      if (runif(1) < lambda/(lambda+mu)) {
        state = c(state, current_state+1)
      } else {
        state = c(state, current_state-1)
      }
    }
  }
  
  state_table = rep(0, 81)
  for (j in 1:(length(time)-1)) {
    state_table[state[j]+1] = state_table[state[j]+1] + time[j+1] - time[j]
  }

  pi = state_table/time[length(state)-1]
  L = sum((0:80)*pi)
  waiting_times[i] = L/lambda
}
```

Construct the CI.
```{r}
print(t.test(waiting_times)$conf.int)
```


# f)
Plot the two functions in one figure.
```{r}
# Setup the grid and define the functions
grid = seq(0, 1, le = 1000)
W_U = 1/(6-5*grid)
W_N = 6/(6-5*grid)

plot(grid, W_U, xlim = c(0, 1), ylim = c(0, 6*max(W_U)),
     xlab = "Value of p", ylab = "Expected time in the UCC (hours)",
     main = expression("W"['U']*"  and W"['N']*"  as a function of p"),
     col = "red", cex.axis = 1.5, cex.lab = 1.5, type = "l", lwd = 4)
lines(grid, W_N, type = "l", lwd = 2, col = "green")
legend("topleft", legend = c(expression("W"['U']), expression("W"['N'])),
       col = c("red", "green"), lty = 1, lwd = 4, cex = 1.5)
```


# g)
Prepare the parameters and the initial conditions of the queueing system.
```{r}
lambda = 5
mu = 6
p = 0.8
length = 50*24

state_normal = c(0)
state_urgent = c(0)
time = c(0)

```

Simulate the birth-and-death processes
```{r}
set.seed(1)
while (tail(time,1) < length) {
  current_state_normal = tail(state_normal,1)
  current_state_urgent = tail(state_urgent,1)
  if ((current_state_normal + current_state_urgent) == 0) {
    # Case: No patients in the UCC
    time = c(time, tail(time,1) + rexp(n=1, rate=lambda))
    if (runif(1) < p) {
      # Subcase: Next arriving patient is of type urgent
      state_urgent = c(state_urgent, 1)
      state_normal = c(state_normal, 0)
    } else {
      # Subcase: Next arriving patient is of type normal
      state_urgent = c(state_urgent, 0)
      state_normal = c(state_normal, 1)
    }
  } else {
    # Case: There is already at least one patient in the UCC
    time = c(time, tail(time,1) + rexp(n=1, rate=lambda+mu))
    if (runif(1) < lambda/(lambda+mu)) {
      # Subcase: The next patient arrives before the treatment is finished
      if (runif(1) < p) {
        # Sub-subcase: The arriving patient is of type urgent
        state_urgent = c(state_urgent, current_state_urgent+1)
        state_normal = c(state_normal, current_state_normal)
      } else {
        # Sub-subcase: The arriving patient is of type normal
        state_urgent = c(state_urgent, current_state_urgent)
        state_normal = c(state_normal, current_state_normal+1)
      }
    } else {
      # Subcase: The treatment gets finished
      if (current_state_urgent > 0) {
        # Sub-subcase: The treatment was on a patient of type urgent
        state_urgent = c(state_urgent, current_state_urgent-1)
        state_normal = c(state_normal, current_state_normal)
      } else {
        # Sub-subcase: The treatment was on a patient of type normal
        state_urgent = c(state_urgent, 0)
        state_normal = c(state_normal, current_state_normal-1)
      }
    }
  }
}
```

Calculate the expected waiting time for a normal/an urgent patient.
```{r}
# The expected time spent in each state
state_table_urgent = rep(0, 81)
state_table_normal = rep(0, 81)
time_urgent = 0
time_normal = 0
for (i in 1:(length(time)-1)) {
  state_table_urgent[state_urgent[i]+1] =
    state_table_urgent[state_urgent[i]+1] + time[i+1] - time[i]
  state_table_normal[state_normal[i]+1] =
    state_table_normal[state_normal[i]+1] + time[i+1] - time[i]
  # Limit the time basis for the calculation for each state
  if (state_urgent[i+1] != state_urgent[i]) {
    time_urgent = time_urgent + time[i+1] - time[i]
  } else {
    time_normal = time_normal + time[i+1] - time[i]
  }
}

# The expected number of patients in the UCC
pi_urgent = state_table_urgent/time_urgent
L_urgent = sum((0:80)*pi_urgent)
pi_normal = state_table_normal/time_normal
L_normal = sum((0:80)*pi_normal)

# The expected waiting time
W_urgent = L_urgent/lambda
W_normal = L_normal/lambda
print(W_urgent)
print(W_normal)
```

Plot both processes in the same figure for the time 0-12 hours.
```{r}
plot(NULL, NULL, xlim = c(0, 12), ylim = c(0, 10),
     xlab = "Time (hours)", ylab = "Number of patients in the UCC",
     main = "Simulation of number of normal and urgent patients in the UCC",
     cex.axis = 1.5, cex.lab = 1.5)
for (i in 1:sum(time < 12)) {
  lines(time[i:(i+1)], rep(state_urgent[i], 2), lwd = 2, col = "red")
}
for (i in 1:sum(time < 12)) {
  lines(time[i:(i+1)], rep(state_normal[i], 2), lwd = 2, col = "green")
}
legend("topleft", legend = c("Urgent patients", "Normal patients"),
       col = c("red", "green"), lty = 1, lwd = 2)
```

Repeat the simulation 30 times.
```{r}
waiting_times_urgent = rep(0, 30)
waiting_times_normal = rep(0, 30)
set.seed(1)
for (i in 1:30) {
  state_urgent = c(0)
  state_normal = c(0)
  time = c(0)
  
  while (tail(time,1) < length) {
    current_state_normal = tail(state_normal,1)
    current_state_urgent = tail(state_urgent,1)
    # Cases if currently there are zero patients or at least one patient
    # Then, append the state and the time stamp of the process
    if ((current_state_normal + current_state_urgent) == 0) {
      time = c(time, tail(time,1) + rexp(n=1, rate=lambda))
      if (runif(1) < p) {
        state_urgent = c(state_urgent, 1)
        state_normal = c(state_normal, 0)
      } else {
        state_urgent = c(state_urgent, 0)
        state_normal = c(state_normal, 1)
      }
    } else {
      time = c(time, tail(time,1) + rexp(n=1, rate=lambda+mu))
      if (runif(1) < lambda/(lambda+mu)) {
        if (runif(1) < p) {
          state_urgent = c(state_urgent, current_state_urgent+1)
          state_normal = c(state_normal, current_state_normal)
        } else {
          state_urgent = c(state_urgent, current_state_urgent)
          state_normal = c(state_normal, current_state_normal+1)
        }
      } else {
        if (current_state_urgent > 0) {
          state_urgent = c(state_urgent, current_state_urgent-1)
          state_normal = c(state_normal, current_state_normal)
        } else {
          state_urgent = c(state_urgent, 0)
          state_normal = c(state_normal, current_state_normal-1)
        }
      }
    }
  }
  
  state_table = rep(0, 81)
  for (j in 1:(length(time)-1)) {
    state_table[state[j]+1] = state_table[state[j]+1] + time[j+1] - time[j]
  }
  
  state_table_urgent = rep(0, 81)
  state_table_normal = rep(0, 81)
  time_urgent = 0
  time_normal = 0
  for (j in 1:(length(time)-1)) {
    state_table_urgent[state_urgent[j]+1] =
      state_table_urgent[state_urgent[j]+1] + time[j+1] - time[j]
    state_table_normal[state_normal[j]+1] =
      state_table_normal[state_normal[j]+1] + time[j+1] - time[j]
    if (state_urgent[j+1] != state_urgent[j]) {
      time_urgent = time_urgent + time[j+1] - time[j]
    } else {
      time_normal = time_normal + time[j+1] - time[j]
    }
  }
  
  pi_urgent = state_table_urgent/time_urgent
  L_urgent = sum((0:80)*pi_urgent)
  pi_normal = state_table_normal/time_normal
  L_normal = sum((0:80)*pi_normal)
  
  waiting_times_urgent[i] = L_urgent/lambda
  waiting_times_normal[i] = L_normal/lambda
}
```

Construct the CIs.
```{r}
print(t.test(waiting_times_urgent)$conf.int)
print(t.test(waiting_times_normal)$conf.int)
```










## Exercise 2
# a)
Set parameters.
```{r}
set.seed(30000)
variance = 0.5^2
mean = 0.5
mean_given <- rep(mean, 5)
mean_grid <- rep(mean, 51)
# Theta values for created grid and given values
theta_given <- c(0.3, 0.35, 0.39, 0.41, 0.45)
theta_grid <- seq(0.25, 0.5, by = 0.005)
# Correlation
correlation <- function(grid_point, given_point){
  exp(-15 * abs(grid_point - given_point)) * (1 + 15 * abs(grid_point - given_point))
}
# Covariance matrix function calculated by: Cov(X,Y) = Corr(X,Y) sqrt(Var(X)Var(Y))
covariance <- function(theta, variance){
  len <- length(theta) 
  cov <- diag(variance, len) # 1's in diagonal
  # Calculation of every value in covariance matrix 
  for (i in 1 : (len - 1)){ 
    for (j in (i + 1) : len){
      correl = correlation(theta[i], theta[j]) 
      cov[i, j] = correl * variance 
      cov[j, i] = correl * variance 
    }
  }
  return(cov)
}

```

Calculations.
```{r}
# Quadratic covariance matrices for grid and given values
cov_given <- covariance(theta_given, variance)
cov_grid <- covariance(theta_grid, variance)

# Creation of Covariance ab, because cov_(grid&given) = cov_(given&grid) ^ T(transposed)
cov_both <- matrix(0, 51, 5)

# Calculation of covariance of grid and given
for (i in 1 : 51){
  for (j in 1 : 5){ 
    cov_both[i, j] = correlation(theta_grid[i], theta_given[j]) * variance 
  }
}

given_values <- c(0.5, 0.32, 0.40, 0.35, 0.60) 

# Using formulas calculated, shown in LateX
total_mean <- mean_grid + cov_both %*% solve(cov_given) %*% (given_values - mean_given)

# Total covariance
total_cov <- cov_grid - cov_both %*% solve(cov_given) %*% t(cov_both)

# Value to calculate bounds
z_sigma <- qnorm(0.9) * sqrt(diag(total_cov))

# Upper and lower bounds
lower_bound <- total_mean - z_sigma
upper_bound <- total_mean + z_sigma
```

Plot.
```{r}
# Plotting
plot(theta_grid, total_mean,
    main =" Estimated score with prediction intervals, 5 given points",
    ylab= expression(Y(theta)),
    xlab = expression(paste("Theta, ", theta)), 
    ylim = c(0.1, 1),
    type="l",
    col="red")
    lines(theta_grid, upper_bound, col = "green", lty=2)
    points(theta_given, given_values, col= "black")
    lines(theta_grid, lower_bound, col="purple", lty=2)
    legend("topright",
    c("estimated fit", "upper bound", "lower bound", "observed values"),
    fill=c("red","purple", "green", "black"))
```

# b)
Plot.
```{r}
# Plotting
plot(theta_grid, pnorm(0.3, mean = total_mean, sd = sqrt(diag(total_cov))),
    main="Conditional probabilities, 5 given points",
    ylab = expression(P(Y(theta) < 0.3)), 
     xlab = expression(paste("Theta, ", theta)), 
     type = "b", col = "black")

# Most optimal theta, max value, for Y(θ) < 0.30
opt_theta = theta_grid[which.max(pnorm(0.3, mean = total_mean, sd = sqrt(diag(total_cov))))]
print(paste0("The most optimal theta to observe probability: ", opt_theta))

# Show value in the distribution
abline(v = opt_theta, col="red")
```

# c)
Calculations.
```{r}
# New mean, theta and covariance matrix for b, with added point
theta_given <- c(0.3, 0.35, 0.39, 0.41, 0.45, 0.33) 
mean_given <- rep(mean, 6)
cov_given <- covariance(theta_given, variance)
new_cov_both <- matrix(0, 51, 6)

# Iterating through and calculating new matrix
for (i in 1 : 51){ 
  for (j in 1 : 6){ 
    new_cov_both[i, j] = correlation(theta_grid[i], theta_given[j]) * variance
  }
}

# New given y, with one added value
new_given_values <- c(0.5, 0.32, 0.40, 0.35, 0.60, 0.4) 

# Total mean and covariance
new_total_mean <- mean_grid + new_cov_both %*% solve(cov_given) %*% (new_given_values - mean_given)
new_total_cov <- cov_grid - new_cov_both %*% solve(cov_given) %*% t(new_cov_both)

# Value to calculate bounds
new_z_sigma <- qnorm(0.9) * sqrt(diag(new_total_cov))

# Bounds
new_upper_bound <- new_total_mean + new_z_sigma 
new_lower_bound <- new_total_mean - new_z_sigma
```

Plots.
```{r}
# Estimated score
plot(theta_grid, new_total_mean,
    main =" Estimated score with prediction intervals, 6 given points",
    ylab= expression(Y(theta)),
    xlab = expression(paste("Theta, ", theta)),
    ylim = c(0.1, 1),
    type="l",
    col="red")
    lines(theta_grid, new_upper_bound, col = "green", lty=2)
    points(theta_given, new_given_values, col= "black")
    lines(theta_grid, new_lower_bound, col="purple", lty=2)
    legend("topright",
    c("estimated fit", "upper bound", "lower bound", "observed values"),
    fill=c("red","purple", "green", "black"))

# Conditional probability
plot(theta_grid, pnorm(0.3, mean = new_total_mean, sd = sqrt(diag(new_total_cov))), 
     main = "Conditional probabilities, 6 given points", 
     ylab = expression(P(Y(theta) < 0.3)), 
     xlab = expression(paste("Theta, ", theta)), 
     type = "b", col = "black")

# Most optimal theta, max value, for Y(θ) < 0.30
new_opt_theta = theta_grid[which.max(pnorm(0.3, mean = new_total_mean, sd = sqrt(diag(new_total_cov))))]
print(paste0("The most optimal theta to observe probability: ", new_opt_theta))

# Show value in the distribution
abline(v = new_opt_theta, col="red")
```

